---
title: "Toward a Multi-Purpose Agent-Based Model for the Brazilian Economy: New Model"
author: "Talitha Speranza"
date: '`r Sys.Date()`'
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_html: default
link-citations: yes
subtitle: New Model
header-includes:
 - \usepackage{amsmath}
 - \usepackage{amssymb}
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

```{marginfigure}
$Q_j(t) = p_j(t) \cdot q_j(t)$
```


# 10. Genetic Evolution

Consumption goods firms need to adapt the following parameters:

* $\beta_j$: proportion between inventory and production;
* $\delta_j$: amortization rate;
* $\alpha_j$: proportion between internal funds and loans for financing;
* $\kappa_j$: prices' sensibility to excessive demand.

The genetic selection of these values depends on basically three other parameters:

* $P$: population maximum size;
* $N$: number of other firms from which to copy chromossomes;
* $T$: number of lags to be stored.

A chromossome is defined as $c_f^l = [\beta_f^l, \delta_f^l, \alpha_f^l, \kappa_f^l]$, where $f$ is the firm index and $l$ is the time index. Therefore, $f \in \{j, f_1, f_2,...,f_N\}$ and $l \in \{t-1,t-2,...,t-T\}$. There are two objective functions in the case of these firms: $\Pi(c_f^l)$, realized profits, and $L(c_f^l)$, total value of loans. The first must be maximized, the second, minimized.


![Flowchart of the genetic algorithm agents use to adapt its behaviour.](AG.pdf)

\newpage

```{r bib, eval= F, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```
