---
title: "Toward an Agent-Based Model for the Brazilian Economy"
subtitle: "Data Initialization"
author: "Talitha Speranza"
date: "April 10, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
# Chunck options
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# Libraries
library(kableExtra)
library(dplyr)

# Constants
mln <- 1*10^6 
tsd <- 1*10^3
bln <- 1*10^9
tln <- 1*10^12

# 
disp.params <- data.frame("Description" = character(), "Unit" = character(), 
                          "Value" = numeric(), stringsAsFactors=FALSE)
params <- data.frame("key" = character(), "value" = numeric(), stringsAsFactors = FALSE)
```

# Number of Agents 

```{r echo=F}
n.consumers <- 2000
```

The number of agents of each type in relation to other types will be based on the real proportions we observe in Brazilian time series and official records. The number of consumers, from now on $E$, is fixed in `r n.consumers` and the quantitity of other agents is calculated from this number. We begin by reading data on Brazilian firms and employees^[Brazilian Institute for Statistics and Geography, [Central Registry of Enterprises (2015)](https://sidra.ibge.gov.br/pesquisa/cempre/quadros/brasil/2015 )]. Some registers of this database are shown below (Table \ref{tab:cce}).

```{r echo = F}
# Read data from Cadastro Central de Empresas
dist.firms <- read.csv("cce.csv", stringsAsFactors = F, encoding = "latin1")
dist.firms[dist.firms == "X"] <- 0
dist.firms[,4:7] <- data.frame(lapply(dist.firms[,4:7], as.integer))

# This column will be used later
dist.firms$avg.min.wages <- as.numeric(gsub(",","\\.",dist.firms$avg.min.wages)) 

# Eliminate NAs
dist.firms <- na.omit(dist.firms)

# Filter 
v1 <- filter(dist.firms, (type == "c" | type == "k") & bracket == "Total")
v2 <- filter(dist.firms, (type == "b"))

# Get totals
tot.firms <- sum(v1$units)
tot.empls <- sum(v1$employees)
tot.banks <- sum(v2$units)

# Get subtotals
v11 <- filter(v1, type == "c")
v12 <- filter(v1, type == "k")
tot.c <- sum(v11$units)
tot.k <- sum(v12$units)
```

```{r echo=F}
# Display table
sub.dist.firms <- head(dist.firms[,-2],20)
names(sub.dist.firms) <- c("","Bracket","Units", "Employees","Salaried","Wage Volume", "Avg. Min. Wages")
kableExtra::kable(sub.dist.firms, format = "latex", booktabs = T, caption = "\\label{tab:cce}Twenty first records of IBGE's Central Registry of Enterprises, 2015. First column represents the specialization according to CNAE 2.0. Bracket is in number of employees. Wage volume is in thousands of R\\$.") %>% 
  kable_styling(latex_options = c("scale_down","hold_position")) %>%
  collapse_rows(columns = 1, latex_hline = "major") 
```

According to this dataset, in 2015 Brazil had approximately `r round(tot.firms/mln,1)` million non-financial private firms, which employed around `r round(tot.empls/mln,1)` million workers, and there were about `r round(tot.banks/tsd, 2)` thousand private banks and other financial firms. The number of consumption goods firms was approximately `r round(tot.c/mln,1)` million, whereas the number of capital goods firms was around `r round(tot.k/mln,1)` million. 

```{r echo = F}
unemp <- read.csv2("desemprego.csv", stringsAsFactors = F, sep="\t")
unemp.rate <- round(mean(unemp[grepl("2015",unemp[,1]),2])/100,3)

tot.workers <- tot.empls/(1 - unemp.rate)
```

We need to consult the unemployment rate series to calculate the total number of workers available in 2015. That year, the unemployment rate was `r unemp.rate*100`% in 2015, as disclosed by IBGE's [_Monthly National Household Survey_](https://sidra.ibge.gov.br/tabela/6381), so the total number of workers can be approximated to `r round(tot.workers/mln,2)` million.

Let $F_1$ be the number of consuption goods firms, $F_2$ the number of capital goods firms, and $B$ the number of banks within the model. Then:

```{r echo = F}
n.con.firms <- round(n.consumers*(tot.c/tot.workers))
n.cap.firms <- round(n.consumers*(tot.k/tot.workers))
n.banks <- round(n.consumers*(tot.banks/tot.workers))
n.unemployed <- round(2000*unemp.rate)
n.employed <- n.consumers - n.unemployed
```

\begin{equation}
F_1 \approx E\cdot\left(\frac{`r round(tot.c/mln,1)`}{`r round(tot.workers/mln,1)`}\right) \approx `r n.consumers`\cdot `r round(tot.c/tot.workers, 3)` \approx `r n.con.firms`
\end{equation}

\begin{equation}
F_2 \approx E\cdot \left(\frac{`r round(tot.k/mln,1)`}{`r round(tot.workers/mln,1)`}\right) \approx `r n.consumers`\cdot `r round(tot.k/tot.workers, 4)` \approx `r n.cap.firms`
\end{equation}

\begin{equation}
B \approx E\cdot \left(\frac{`r round(tot.banks/mln,1)`}{`r round(tot.workers/mln,1)`}\right) \approx `r n.consumers`\cdot `r round(tot.banks/tot.workers, 3)` \approx `r n.banks`
\end{equation}

Table \ref{tab:agents} shows a list of the parameters already discussed.

```{r echo = F}
#  Fill 'disp.params'
disp.params[1,] <- list("Number of Consumers", "consumers", n.consumers)
disp.params[2,] <- list("Number of Unemployed Consumers", "consumers", n.unemployed)
disp.params[3,] <- list("Number of Consumption Goods Firms", "firms", n.con.firms)
disp.params[4,] <- list("Number of Capital Goods Firms", "firms", n.cap.firms)
disp.params[5,] <- list("Number of Banks", "banks", n.banks)

params[1,] <- list("n_consumers", n.consumers)
params[2,] <- list("n_unemployed", n.unemployed)
params[3,] <- list("n_con_firms", n.con.firms)
params[4,] <- list("n_cap_firms",n.cap.firms)
params[5,] <- list("n_banks",n.banks)

knitr::kable(disp.params, caption="\\label{tab:agents}Number of agents of each type.", row.names = F, booktabs = T)
```

## Gross Domestic Product

```{r echo = F}
real.gdp = 1817242668600
nom.gdp = 6000572000000
gdp.per.worker = round(real.gdp/tot.workers)
gdp = round(gdp.per.worker*n.consumers)
gdp.per.empl = round(gdp/n.employed)
#min.wage = round((788*gdp/nom.gdp),2) 
```

The initial Gross Domestic Product (GDP) of the model was defined in terms of the population of consumers, just as every other model parameter. First, we collect the values of the real^[World Bank, [World Development Indicators](http://databank.worldbank.org/data/reports.aspx?source=2&series=NY.GDP.MKTP.KD.ZG&country=BRA#) (2017)]
and the nominal^[Central Bank of Brasil (BCB), [Department of Economics](http://www.bcb.gov.br/pec/Indeco/Port/indeco.asp)] GDP. In 2015, real GDP was around R\$ `r round(real.gdp/tln,1)` trillion. Nominal GDP, provided by the [Department of Economics](http://www.bcb.gov.br/pec/Indeco/Port/indeco.asp) of the Central Bank of Brasil (BCB/Depec), was about R\$ `r round(nom.gdp/tln,1)` trillion in 2015. 

If $y$ is the real GDP per worker in 2015, then the model's initial GDP, $Y$, can be written as

\begin{equation}
Y = y \cdot E \approx \left(\frac{`r formatC(round(real.gdp/mln,1),digits=2)`}{`r round(tot.workers/mln,1)`}\right) \cdot `r n.consumers` \approx (`r formatC(gdp.per.worker, digits=2)`) \cdot `r n.consumers` \approx `r formatC(gdp,digits = 2)` 
\end{equation}

```{r echo = F }
# Percentage of salaries in GDP 
perc.sal = round(2516592 / 5778953,2)
gross.sal = perc.sal*gdp
gross.profit = gdp - gross.sal
prf.per.empl = round(gross.profit/n.employed)
```

At this point we also need to obtain the percentage of salaries in GDP^[The GDP can be defined as the total sum of salaries and profits in an economy]. The most recently published data on detailed composition of Brazilian GDP are from 2014's [_National Accounts System_](https://ww2.ibge.gov.br/home/estatistica/economia/contasnacionais/2014/defaulttab_xls.shtm) (IBGE), but these are surely good approximations for 2015's values. In 2014, wages accounted for `r round(100*perc.sal, 1)`\% of the GDP. Hence, profits were `r round(100*(1-perc.sal), 1)`\% of the GPD that same year. These figures allowed us to set some other parameters, as shown in table \ref{tab:gdp}.

```{r echo = F}
disp.params[6,] <- list("Real GDP", "goods", as.integer(gdp))
disp.params[7,] <- list("Real GDP per employed worker (last 12 months)", "goods", round(gdp.per.empl,2))
disp.params[8,] <- list("Percentage of salaries in GDP","percentage", as.integer(round(100*perc.sal,2)))
disp.params[9,] <- list("Profit per employed worker (last 12 months)", "goods", round(prf.per.empl,2))

params[6,] <- list("gdp", gdp)
params[7,] <- list("gdp_per_empl", round(gdp.per.empl/12,2))
params[8,] <- list("perc_sal", perc.sal)
params[9,] <- list("prf_per_empl",round(prf.per.empl/12,2))

kable(disp.params[6:9,-4], caption="\\label{tab:gdp}GDP and its components.", row.names = F)
```

Initial profits were assigned to each firm according to its number of workers. The unit of profit, $\pi_u$, is shown in the table above (*Profit per employed worker*). If a firm has $N$ employees, its starting profit is $N \cdot \pi_u \cdot x$, where $x \sim U(0,2)$. This way, we introduce more heterogeneity into the system, by allowing some firms to begin with littler or no profits and other with outstanding past results. Given that the mean of $x$ is 1, the average initial profit per worker matches real data.

```{r echo = F}
capc <- read.csv2("used_capacity.csv", stringsAsFactors = F, sep="\t", encoding = "latin1")
capc <- round(mean(capc[grepl("2015",capc[,1]),2])/100,3)

params[10,] <- list("used_cap",capc)
```


```{r echo = F}
#knitr::kable(params)
```

## Number of Employees Per Firm

```{r echo=F}
get.dist <- function(tp, n.firms, dist.firms){
  
  filtered <- filter(dist.firms, type == tp, bracket != "Total")[,3:5]
  filtered <- aggregate(filtered[,-1], by = list(filtered$bracket), FUN = sum)
  
  s = sum(filtered$units)
  filtered$units = ceiling((filtered$units/s)*n.firms)
  filtered$employees = floor((filtered$employees/tot.empls)*n.employed)
  
  dist <- data.frame("n.firms" = integer(), "n.emps" = integer(), "tot.emps" = integer())
  j = 1
  
  for(i in 1:nrow(filtered)){
    group <- filtered[i,]
    emps.per.firm <- floor(group$employees/group$units)
    dif <- group$employees - (emps.per.firm*group$units)
    
    if(dif > 0){
      dist[j,] <- list(dif, emps.per.firm + 1, dif*(emps.per.firm + 1))
      j = j + 1
    }
    
    dist[j,] <- list(group$units - dif, emps.per.firm, (group$units - dif)*emps.per.firm)
    j = j + 1
  }
  
  agg.dist <- aggregate(dist, list(dist$n.emps),sum)[,-1]
  return(agg.dist)
}

dist.con <- get.dist("c",n.con.firms, dist.firms)
dist.cap <- get.dist("k",n.cap.firms, dist.firms)
n.emps <- sum(dist.con$n.firms*dist.con$n.emps) + sum(dist.cap$n.firms*dist.cap$n.emps)
```

The *Central Register of Enterprises* also serves the purpose of calculating how many employees each firm must have. Before all else, we extract two groups from the dataset: consumption goods firms and capital goods firms. Then, we aggregate data by brackets, adding grouped columns for each subset. For instance, capital goods subset would look like the table shown below (table \ref{tab:caps}).

```{r echo = F}
filtered <- filter(dist.firms, type == "k", bracket != "Total")[,3:5]
filtered <- aggregate(filtered[,-1], by = list(filtered$bracket), FUN = sum)
names(filtered) <- c("Bracket","Units","Employees")
knitr::kable(filtered, caption = "\\label{tab:caps}Aggregated data for capital goods firms, by brackets. Some columns are not shown, because they are not going to be useful in the following calculations.") 
```

The next step is to divide each element from the column *units* by the sum of all elments, so as to find the percentages they represent. On the other hand, elements of *employees* are divided by `r tot.empls`, which is the total number of workers registered in the original dataset (the sum of employees from capital and consumption goods firms). This way, this column will represent percentages of the total number of workers. Having these columns of percentages, we multiply the first by the number of the associated type of firm within the model (`r n.cap.firms` capital goods firms and `r n.con.firms` consumption goods firms) and the second by the number of employed consumers (`r n.employed`). The results are presented in table \ref{tab:dist.percs}.

```{r echo = F}
s = sum(filtered$Units)
filtered$units = ceiling((filtered$Units/s)*n.cap.firms)
filtered$employees = ceiling((filtered$Employees/tot.empls)*n.employed)
knitr::kable(filtered, caption="\\label{tab:dist.percs} Number of firms and employees per bracket.")
```

Note that the second column, *employees* represent the total number of employees in all units. Therefore, we find the average number of employees per firm and round it down to the nearest integer. Some firms will have this number of employees, while others will have one more, in order to absorb the remainder. For instance, if *employees* is 535 and *units* is 100, 65 firms will have 5 employees and 35 firms, 6 employees, since $\lceil 535/100 \rceil = 5$ and the remainder is 35, which imposes that 35 firms have $5 + 1 = 6$ employees. We now have the initial distribution of employees per firm. The result of such calculations are shown in table \ref{tab:dist.cap}. 

```{r echo = F}
temp1 <- cbind(dist.con[,-3],"c")
temp2 <- cbind(dist.cap[,-3],"k")
names(temp1)[3] <- "type"
names(temp2)[3] <- "type"

dist.fir <- rbind(temp1,temp2)

dist.fir[2,1] = 100 

sum = 0

for(i in 1:nrow(dist.fir)){
  sum = sum + dist.fir[i,1]*dist.fir[i,2]
}

params[2,] <- list("n_unemployed", n.consumers - sum)

write.csv(dist.fir, file = "/home/talithafs/Dropbox/Dissertação/Modelo/abm/input/dist_fir.csv", row.names = F)

names(dist.cap) <- c("N. Firms", "N. Employees", "Tot. Employees")
names(dist.con) <- names(dist.cap)

kableExtra::kable(list(dist.cap,dist.con), caption = "\\label{tab:dist.cap}Distribution of employees per firm. The first table presents results for capital goods firms. The second, for consumption goods firms.", format = "latex", booktabs = T) %>%
kable_styling(font_size = 8)
```

## Distribution of Wealth and Income 

Workers belong to social classes that are defined by income brackets. Each of these brackets concentrates a certain amount of wealth and is delimited by monthly earnings in terms of minimum wages. In this session we show how we attributed a social class to every worker and set his income and wealth accordingly. Data on the distribution of wealth and income are yearly disclosed by the [Federal Revenue Office](https://idg.receita.fazenda.gov.br/dados/receitadata/estudos-e-tributarios-e-aduaneiros/estudos-e-estatisticas/11-08-2014-grandes-numeros-dirpf/grandes-numeros-dirpf-capa). The dataset is presented below (table \ref{tab:income}).

```{r echo=F}
# Read data
dist.inc <- read.csv("distribuicao_renda_riqueza.csv", stringsAsFactors = F, encoding = "latin1")
dist.inc[,-1] <- data.frame(lapply(dist.inc[,-1], function(x) {
  y <- gsub(',', '', x)
  as.numeric(y)
}
))

names(dist.inc) = c("faixa","pessoas","renda","rend.trib.ex","rend.ist","bens")
dist.inc[,-1] <- data.frame(lapply(dist.inc[,-1], as.integer))

disp <- dist.inc
names(disp) <- c("Bracket","Population","Income","Not Exempt","Exempt","Wealth")

knitr::kable(disp, caption = "\\label{tab:income}Income and wealth distribution.")
```

```{r echo=F}
# Get total wealth
wealth = (dist.inc[nrow(dist.inc),ncol(dist.inc)])*mln

# Find wealth in terms of the model's GDP
wealth = (wealth/nom.gdp)*gdp
```

Wealth is on the last column of this dataset, and totals on the last row. Hence, total wealth is element (`r c(nrow(dist.inc),ncol(dist.inc))`) of table \ref{tab:income} multiplied by 1 million (data are in R\$ millions). To find the model's total wealth proportional to its GDP, we have to divide it by the nominal GDP, since income and wealth are in nominal values, and then multiply it by the model's initial GDP. By doing so, we find that the model initial total wealth is `r round(wealth/mln,1)` million.

```{r echo=F}
# Collapse "renda isenta" and "renda não isenta" 
dist.inc$renda <- (dist.inc$renda + dist.inc$rend.ist + dist.inc$rend.trib.ex)*mln
dist.inc <- dist.inc[,c(-4,-5)]
dist.inc$bens = dist.inc$bens*mln

# Calculate totals, column-wise
tot <- dist.inc[nrow(dist.inc),]
```

Given that the table separates income that is exempt from taxes (column 5) from income that is not exempt (column 3), we collapse the two values by adding them. Column 4 is removed because it contains income from investments and real-state, but the current version of the model does not incorporate these.

[IBGE](https://thiagorodrigo.com.br/artigo/faixas-salariais-classe-social-abep-ibge/) divides the population in social classes according to the number of minimum salaries workers earn each month. So we aggregate the lines of table \ref{tab:income} following this classsification.

```{r echo=F}
# Aggregate by social class 
filt <- dist.inc$faixa %in% c("Até 1/2","Mais de 1/2 a 1", "Mais de 1 a 2")
class.E = dist.inc[filt,]
class.E = data.frame(lapply(class.E[,-1], sum))
class.E$faixa = "E - Between 0+ and 2 MS"
class.E[,-4] = class.E[,-4] / tot[,-1]

filt <- dist.inc$faixa %in% c("Mais de 2 a 3","Mais de 3 a 5")
class.D = dist.inc[filt,]
class.D = data.frame(lapply(class.D[,-1], sum))
class.D$faixa = "D - Between 2+ and 5 MS"
class.D[,-4] = class.D[,-4] / tot[,-1]

filt <- dist.inc$faixa %in% c("Mais de 5 a 7","Mais de 7 a 10")
class.C = dist.inc[filt,]
class.C = data.frame(lapply(class.C[,-1], sum))
class.C$faixa = "C - Between 5+ and 10 MS"
class.C[,-4] = class.C[,-4] / tot[,-1]

filt <- dist.inc$faixa %in% c("Mais de 10 a 15","Mais de 15 a 20")
class.B = dist.inc[filt,]
class.B = data.frame(lapply(class.B[,-1], sum))
class.B$faixa = "B - Between 10+ and 20 MS"
class.B[,-4] = class.B[,-4] / tot[,-1]

filt <- dist.inc$faixa %in% unique(dist.inc$faixa)[10:17]
class.A = dist.inc[filt,]
class.A = data.frame(lapply(class.A[,-1], sum))
class.A$faixa = "A - More than 20 MS"
class.A[,-4] = class.A[,-4] / tot[,-1]

props <- rbind(class.A, class.B, class.C, class.D, class.E)
classes <- c("A","B","C","D","E")

vec.emps <- round(props$pessoas*n.employed)
vec.unemps <- round(props$pessoas*n.unemployed)

# disp.params <- data.frame(Variable = "Number of Consumers",
#                   Class = classes,
#                   Unit = "employed workers - unemployed workers",
#                   "Value 1" = vec.emps,
#                   "Value 2" = vec.unemps)

#params[10:14,] <- list(paste0("n_emps_class_",classes), vec.emps)
#params[15:19,] <- list(paste0("n_unemps_class_",classes), vec.unemps)

vec.wages <- round((props$renda*gross.sal/round(props$pessoas*n.employed))/12)

# df <- data.frame(Variable = "Monthly wage of an employed consumer",
#                  Class = classes,
#                  Unit = "goods",
#                  "Value 1" = vec.wages,
#                  "Value 2" = "-")
# disp.params <- rbind(disp.params, df)

#params[20:24,] <- list(paste0("mon_wage_class_",classes),vec.wages)

vec.wealth <- round((props$bens*wealth)/(props$pessoas*n.consumers))

# df <- data.frame(Variable = "Wealth of a consumer",
#                  Class = classes,
#                  Unit = "goods",
#                  "Value 1" = vec.wealth,
#                  "Value 2" = "-")
# disp.params <- rbind(disp.params, df)

#params[25:29,] <- list(paste0("wealth_class_",classes),vec.wealth)

```

```{r echo=F}
# NAO FALEI DISSO!

filt <- dist.inc$faixa %in% c("Até 1/2","Mais de 1/2 a 1", "Mais de 1 a 2")
class.E = dist.inc[filt,]

avg.mon.inc = sum(class.E$renda/12)/sum(class.E$pessoas) 
n.min = avg.mon.inc / 788 #esse era o sal min em 2015
min.wage = n.min*vec.wages[5]
params[nrow(params)+1,] <- list("min_wage",round(min.wage,2))
```

```{r echo = F}
dist.inc <- data.frame(class = classes, emps = vec.emps, unemps = vec.unemps, mon.wage = vec.wages, wealth = vec.wealth)
write.csv(dist.inc, "/home/talithafs/Dropbox/Dissertação/Modelo/abm/input/dist_inc.csv", row.names = F)

disp.params <- dist.inc[,-1]
names(disp.params) <- c("Employed","Unemployed","Monthly Wage","Wealth")
row.names(disp.params) <- classes 

# kableExtra::kable(disp.params, format = "latex", booktabs = T, caption = "\\label{tab:inc.params} Wealth and income of our model's consumers.") %>% 
#   kable_styling(latex_options = c("scale_down","hold_position")) %>%
#   collapse_rows(columns = 1, latex_hline = "major") 
kableExtra::kable(disp.params, format = "latex", booktabs = T, caption = "\\label{tab:inc.params} Wealth and income of our model's consumers.")  %>% 
   kable_styling(latex_options = "hold_position")
```


First, we found the percentage of people in each class in the Brazilian population. Then, we applied those percentages to the model and, using the unemployment rate, calculated the numbers of employed and unemployed workers within a class. For instance, 8.45\% of the Brazilian population belongs to class A. Hence, 169 ($= 2000\cdot0.845$) workers in our model must belong to this class, 14 ($= 169\cdot0.83$) being initially unemployed and 155 ($= 169 - 14$), employed. Results of these computations are exhibited above (table \ref{tab:inc.params}).

## Debt

Non-corporate debt is expressed as a percentage of household income^[BCB, Household Debt without Mortgage Loans (Series 20400)]. Since families in our model are composed of one member only, this percentage represents approximately how much debt each worker has in relation to his income, in the first run. Actually, non-corporate debt rate was assigned to consumers following a $U(0,54.34)$ distribution, whose mean is exactly the percentage found in Brazilian data ($27.17$, as shown in the table below). We did so because some individuals might not hold debt at all, but others may be severely indebted. 

```{r echo = F}
hh.debt <- read.csv2("divida_familias.csv",stringsAsFactors = F)
hh.debt <- round(mean(as.numeric(hh.debt[grepl("2015",hh.debt$Date),2])),2)

f.debt <- read.csv2("divida_empresas.csv",stringsAsFactors = F)
f.debt[,2] <- gsub(",","",f.debt[,2])
f.debt <- round(mean(as.numeric(f.debt[grepl("2015",f.debt$Date),2])),2)

prop.f.debt <- round(f.debt*1000000/nom.gdp,2)
debt.per.empl <- round(prop.f.debt*gdp/n.employed,2)

disp.params <- data.frame(Description = character(), Unit = character(), Value = numeric(), stringsAsFactors=FALSE)
disp.params[1,] <- list("Mean household debt (% last 12 months earnings)","goods", hh.debt)
disp.params[2,] <- list("Mean firms debt (% GDP)", "goods", 100*prop.f.debt)
disp.params[3,]<- list("Firms debt per employee", "goods", debt.per.empl)

knitr::kable(disp.params, caption = "\\label{tab:debt}Debt parameters", row.names = F)
```

Corporate debt^[BCB, Credit operations outstanding by type of borrower - Private sector (Series 22047)] is split among firms roughly in proportion to the their number of employees. It means that the whole amount of debt is divided into the total number of workers, and the result of this division is the unit of debt, which is then multiplied by a uniform random variable between 0 and 2 (such that the mean is exactly the unit of debt). For instance, a firm with 5 workers has an initial debt of 5 times this randomized unit. Again, the idea of using a random element is to introduce heterogeneity and account for different levels of indebtness. This level, however, refer only to the long-term debt. We will talk about short-term debt later.

As usual, real values were those of 2015 and compatibilization with model units was achieved through simple proportions.

```{r echo=F}
params[nrow(params)+1,] <- list("hh_debt",round(hh.debt/100, 2))
params[nrow(params)+1,] <- list("debt_per_empl",debt.per.empl)

# NAO FALEI SOBRE ISSO!!
params[nrow(params)+1,] <- list("saving_perc",0.164)
```

```{r echo = F}
params[nrow(params)+1,] <- list("gross_sal",gross.sal/12)
params[nrow(params)+1,] <- list("debt_to_equity", 0.04)
```

## Interest Rates

Interest rates charged on loans are different across types of agents. The risk of a worker defaulting on his debt is greater than the risk of a firm going bankrupt. Therefore, a worker must pay higher interests on loans. The same logic applies to firms of different sizes. Smaller firms offers more risk, so they must be charged higher interests. 

In view of the lack of data on the dimensions of the differences among these charges, interest rates will be calibrated along with several other parameters in the second model. However, in the first model, banks choose initial values for each type of borrower obeying a very simple rule. A mean rate, which is the interest demanded from medium firms, is attributed to each bank accoring to a normal distribution, whose mean is the real mean of interest rates on new credit operations^[BCB-Dstat, Monthly average interest rate of nonearmarked new credit operations - Non-financial corporations (Series 25437) and Households (Series 25462)] in 2015. Banks add or subtract a uniformly random value between 0 and 0.5 to set other interest rates. For example, if the interest charged to medium firms is 2.08, then to small firms it could be 2.29 ($= 2.08 + 0.21$), to micro firms, 2.43 ($= 2.29 + 0.14$), and to large firms, 1.65 ($= 2.08 - 0.43$).  

```{r echo = F}
interest <- read.csv2("taxa_de_juros.csv",stringsAsFactors = F)
f.ints <- round(mean(as.numeric(interest[grepl("2015",interest$Date),2])),2)
h.ints <- round(mean(as.numeric(interest[grepl("2015",interest$Date),3])),2)

s.ints <- read.csv("poupanca.csv", stringsAsFactors = F, encoding = "latin1")
s.ints <- mean(as.numeric(s.ints[grepl("2015",s.ints[,1]),2]))

require(BETS)
selic <- BETSget(4390)
mean.selic <- mean(window(selic, start=c(2015,1), end=c(2015,12)))


disp.params[1,] <- list("Mean interest rate on loans - workers", "monthly yield", h.ints)
disp.params[2,] <- list("Mean interest rate on loans - firms", "monthly yield", f.ints)
disp.params[3,] <- list("Mean interest rate on savings", "monthly yield", round(s.ints,2))
disp.params[4,] <- list("Bankruptcy rate", "yearly %", 3.1)
disp.params[5,] <- list("Mean SELIC rate", "monthly", mean.selic)

params[nrow(params)+1,] <- list("mean_int_cons",round(h.ints/100,5))
params[nrow(params)+1,] <- list("mean_int_firms",round(f.ints/100,5))
params[nrow(params)+1,] <- list("mean_savs_int",round(s.ints/100,5))
params[nrow(params)+1,] <- list("bkr_rate",0.031)
params[nrow(params)+1,] <- list("mean_selic",round(mean.selic,3))

write.csv(params, file = "/home/talithafs/Dropbox/Dissertação/Modelo/abm/input/parameters.csv",row.names = F)

knitr::kable(disp.params[1:4,], caption = "\\label{tab:int}Interest rates", row.names = F)
```



